---
title: "Proportion of Deliveries Paid by SHA"
subtitle: "SHA Coverage of Total Deliveries · 2024"
date: today
date-format: "D MMMM YYYY"
author: "Health Services Analytics Team"
format:
  html:
    theme: [flatly, custom.scss]
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    toc-location: left
    number-sections: false
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    fig-width: 10
    fig-height: 5
    fig-dpi: 150
    embed-resources: true
    page-layout: full
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false

# ── Install / load packages ──────────────────────────────────
pkgs <- c("tidyverse", "lubridate", "scales", "glue",
          "plotly", "gt", "gtExtras", "htmltools")
installed <- rownames(installed.packages())
to_install <- pkgs[!pkgs %in% installed]
if (length(to_install)) install.packages(to_install, quiet = TRUE)
invisible(lapply(pkgs, library, character.only = TRUE))

# ── Colour palette ────────────────────────────────────────────
pal <- list(
  bg      = "#F4F6FB",
  card    = "#FFFFFF",
  primary = "#1A56DB",   # Total deliveries (KHIS)
  accent  = "#0E9F6E",   # SHA claimed
  gap     = "#DBEAFE",   # shaded gap fill between lines
  green   = "#0E9F6E",
  amber   = "#D97706",
  red     = "#E02424",
  text    = "#111928",
  muted   = "#6B7280",
  grid    = "#E5E7EB",
  border  = "#D1D5DB"
)

# ── ggplot2 base theme ────────────────────────────────────────
theme_del <- function() {
  theme_minimal(base_family = "sans") +
    theme(
      plot.background  = element_rect(fill = pal$bg,   colour = NA),
      panel.background = element_rect(fill = pal$bg,   colour = NA),
      panel.grid.major = element_line(colour = pal$grid, linewidth = 0.4),
      panel.grid.minor = element_blank(),
      axis.text        = element_text(colour = pal$muted, size = 9),
      axis.title       = element_text(colour = pal$text,  size = 10),
      plot.title       = element_text(colour = pal$text,  size = 13,
                                      face = "bold", margin = margin(b = 4)),
      plot.subtitle    = element_text(colour = pal$muted, size = 9,
                                      margin = margin(b = 12)),
      legend.position  = "bottom",
      legend.text      = element_text(colour = pal$muted, size = 8),
      plot.margin      = margin(12, 16, 12, 16),
      strip.text       = element_text(colour = pal$text, face = "bold", size = 10)
    )
}
theme_set(theme_del())
```

```{r}
#| label: data-generation
#| include: false

# ══════════════════════════════════════════════════════════════
# MOCKUP: All data below is DUMMY / SYNTHETIC.
# Backend team should replace `df_sha` and `df_khis` with real
# queries per the spec below.
# ══════════════════════════════════════════════════════════════

set.seed(42)

# ── BACKEND SPEC – SHA claims ────────────────────────────────
# Source : claims / encounters table
# Filter : intervention_code IN ('SHA-08-005',   -- Vaginal delivery
#                                'SHA-08-006',   -- Caesarean section
#                                'SHA-08-007')   -- Multiple births
# Columns:
#   county        – county of facility
#   claim_date    – date of service (used to derive month)
#   claim_count   – COUNT(*) per county-month
# Join key: county + floor_date(claim_date, "month")
# ─────────────────────────────────────────────────────────────

# ── BACKEND SPEC – KHIS total deliveries ─────────────────────
# Source    : DHIS2 API  https://hiskenya.dha.go.ke/
# Element   : BDsWWkJHuce  (Total deliveries)
# Pull by   : month, orgUnit = county
# Columns:
#   county       – county name (must match SHA county labels)
#   period_month – DHIS2 period converted to first-of-month Date
#   khis_count   – total deliveries reported in DHIS2
# Note: pull monthly; join on county + period_month
# ─────────────────────────────────────────────────────────────

counties <- c("Nairobi","Mombasa","Kisumu","Nakuru","Eldoret",
               "Nyeri","Meru","Thika","Kakamega","Kisii",
               "Machakos","Garissa","Kitui","Bungoma","Embu",
               "Kilifi","Siaya","Bomet","Nandi","Vihiga")

months_seq <- seq(as.Date("2024-01-01"), as.Date("2024-11-01"), by = "month")

# SHA monthly claimed deliveries per county
df_sha <- expand_grid(county = counties, period_month = months_seq) %>%
  mutate(
    sha_count = round(rnorm(n(), mean = 320, sd = 90)) %>% pmax(20)
  )

# KHIS monthly total deliveries per county (always >= sha_count on average)
df_khis <- expand_grid(county = counties, period_month = months_seq) %>%
  mutate(
    khis_count_raw = round(df_sha$sha_count / runif(n(), 0.45, 0.95)) %>% pmax(30)
  )

# ── OUTLIER FLAGGING & WINSORISATION (IQR method) ───────────
# BACKEND NOTE: Apply this logic server-side before serving data,
# or replicate in the frontend data-prep layer.
# Flag a county-month KHIS value as outlier if it falls outside
# [Q1 - 1.5*IQR, Q3 + 1.5*IQR] computed across all county-months.
# Replace flagged values with the nearest fence (lower or upper).

iqr_bounds <- function(x) {
  q1  <- quantile(x, 0.25, na.rm = TRUE)
  q3  <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  list(lower = q1 - 1.5 * iqr, upper = q3 + 1.5 * iqr)
}

bounds <- iqr_bounds(df_khis$khis_count_raw)

df_khis <- df_khis %>%
  mutate(
    is_outlier  = khis_count_raw < bounds$lower | khis_count_raw > bounds$upper,
    khis_count  = case_when(
      khis_count_raw < bounds$lower ~ bounds$lower,
      khis_count_raw > bounds$upper ~ bounds$upper,
      TRUE                          ~ khis_count_raw
    ) %>% round()
  )

# ── JOINED DATASET ────────────────────────────────────────────
# BACKEND NOTE: LEFT JOIN khis onto sha so counties with zero SHA
# claims are preserved (per AC2 "include counties with zero values").
df <- df_khis %>%
  left_join(df_sha, by = c("county", "period_month")) %>%
  mutate(
    sha_count  = replace_na(sha_count, 0),
    proportion = sha_count / khis_count,           # N / D
    proportion = pmin(proportion, 1)               # cap at 100 %
  )

# ── GLOBAL FILTER DEFAULTS ────────────────────────────────────
# FRONTEND NOTE: Replace with reactive inputs:
#   input$date_range – dateRangeInput → filters period_month
#   input$county     – selectizeInput (multi) → filters county
# These two filters cascade and affect all three components.

filter_date_min <- as.Date("2024-01-01")    # MOCK → input$date_range[1]
filter_date_max <- as.Date("2024-11-30")    # MOCK → input$date_range[2]
filter_county   <- counties                 # MOCK → input$county (all)

df_f <- df %>%
  filter(
    period_month >= filter_date_min,
    period_month <= filter_date_max,
    county %in% filter_county
  )

# ── SUMMARY CARD METRICS ──────────────────────────────────────
# Reported for the most recent complete month in the filtered window.
cur_month  <- max(df_f$period_month)
prev_month <- cur_month %m-% months(1)

summary_cur  <- df_f %>% filter(period_month == cur_month)  %>%
  summarise(D = sum(khis_count), N = sum(sha_count))
summary_prev <- df_f %>% filter(period_month == prev_month) %>%
  summarise(D = sum(khis_count), N = sum(sha_count))

prop_cur  <- summary_cur$N  / summary_cur$D
prop_prev <- summary_prev$N / summary_prev$D

delta_D    <- summary_cur$D - summary_prev$D
delta_N    <- summary_cur$N - summary_prev$N
delta_prop <- prop_cur - prop_prev

fmt_delta <- function(x, unit = "", good_dir = 1) {
  # good_dir: +1 = higher is better, -1 = lower is better
  colour <- if (x * good_dir > 0) pal$green else if (x * good_dir < 0) pal$red else pal$muted
  sign   <- if (x >= 0) "▲ +" else "▼ "
  val    <- if (unit == "%") glue("{sign}{round(abs(x)*100,1)} pp")
            else             glue("{sign}{format(round(abs(x)), big.mark=',')} {unit}")
  label  <- glue("{val} vs prev month")
  list(label = label, colour = colour)
}

d_D    <- fmt_delta(delta_D,    unit = "deliveries", good_dir =  1)
d_N    <- fmt_delta(delta_N,    unit = "deliveries", good_dir =  1)
d_prop <- fmt_delta(delta_prop, unit = "%",           good_dir =  1)

# ── MONTHLY TREND (national aggregate) ───────────────────────
monthly <- df_f %>%
  group_by(period_month) %>%
  summarise(
    khis_total = sum(khis_count),
    sha_total  = sum(sha_count),
    outliers   = sum(is_outlier),
    .groups    = "drop"
  ) %>%
  mutate(
    proportion = sha_total / khis_total,
    gap        = khis_total - sha_total,
    month_lbl  = format(period_month, "%b %Y")
  )

# ── COUNTY TABLE ──────────────────────────────────────────────
county_tbl <- df_f %>%
  group_by(county) %>%
  summarise(
    khis_count = sum(khis_count),
    sha_count  = sum(sha_count),
    .groups    = "drop"
  ) %>%
  mutate(
    proportion  = sha_count / khis_count,
    prop_pct    = round(proportion * 100, 1),
    prop_colour = case_when(
      proportion >= 0.80 ~ "green",
      proportion >= 0.50 ~ "amber",
      TRUE               ~ "red"
    )
  ) %>%
  arrange(proportion)   # default sort: ascending proportion (AC2)
```

```{r}
#| label: card-css
#| echo: false
#| results: asis

cat('
<style>
  .card-row   { display: flex; gap: 14px; flex-wrap: wrap; margin-bottom: 28px; }
  .metric-card {
    flex: 1 1 160px;
    background: #fff;
    border: 1px solid #E5E7EB;
    border-radius: 10px;
    padding: 18px 20px 14px;
    box-shadow: 0 1px 4px rgba(0,0,0,.06);
  }
  .metric-card .card-label  { font-size: 11px; color: #6B7280; text-transform: uppercase;
                               letter-spacing: .06em; margin-bottom: 6px; }
  .metric-card .card-value  { font-size: 28px; font-weight: 700; color: #111928;
                               line-height: 1.1; }
  .metric-card .card-delta  { font-size: 11px; margin-top: 8px; font-weight: 500; }
  .metric-card .card-sub    { font-size: 10px; color: #9CA3AF; margin-top: 2px; }
  .section-label {
    font-size: 11px; text-transform: uppercase; letter-spacing: .08em;
    color: #6B7280; margin-bottom: 10px; font-weight: 600;
  }
  /* Proportion badge colours */
  .badge-green { background:#D1FAE5; color:#065F46; border-radius:999px;
                 padding:2px 10px; font-size:12px; font-weight:600; }
  .badge-amber { background:#FEF3C7; color:#92400E; border-radius:999px;
                 padding:2px 10px; font-size:12px; font-weight:600; }
  .badge-red   { background:#FEE2E2; color:#991B1B; border-radius:999px;
                 padding:2px 10px; font-size:12px; font-weight:600; }
</style>
')
```

## AC1 · National Summary Cards {#sec-cards}

::: section-label
`r format(cur_month, "%B %Y")` · SHA Claimed Deliveries ÷ KHIS Total Deliveries
:::

```{r}
#| label: metric-cards
#| echo: false
#| results: asis

# ── BACKEND SPEC ──────────────────────────────────────────────
# D (denominator) = SUM(khis_count) for selected period + county
# N (numerator)   = SUM(sha_count)  for selected period + county
# Proportion      = N / D  (expressed as %)
# Delta           = current month metric − previous month metric
# Cards re-render on every filter change.
# ─────────────────────────────────────────────────────────────

card <- function(label, value, delta_label = NULL, delta_colour = NULL, sub = NULL) {
  delta_html <- if (!is.null(delta_label))
    glue('<div class="card-delta" style="color:{delta_colour}">{delta_label}</div>')
  else ""
  sub_html <- if (!is.null(sub))
    glue('<div class="card-sub">{sub}</div>')
  else ""
  glue('
  <div class="metric-card">
    <div class="card-label">{label}</div>
    <div class="card-value">{value}</div>
    {delta_html}
    {sub_html}
  </div>')
}

cards_html <- paste0(
  '<div class="card-row">',

  card(
    label       = "Total Deliveries (D)",
    value       = format(summary_cur$D, big.mark = ","),
    delta_label = d_D$label,
    delta_colour= d_D$colour,
    sub         = "Source: KHIS (BDsWWkJHuce)"
  ),

  card(
    label       = "SHA Claimed Deliveries (N)",
    value       = format(summary_cur$N, big.mark = ","),
    delta_label = d_N$label,
    delta_colour= d_N$colour,
    sub         = "SHA-08-005 / 006 / 007"
  ),

  card(
    label       = "Coverage Proportion (N/D)",
    value       = glue("{round(prop_cur * 100, 1)}%"),
    delta_label = d_prop$label,
    delta_colour= d_prop$colour,
    sub         = "🟢 ≥80%  🟡 50–79%  🔴 <50%"
  ),

  '</div>'
)
cat(cards_html)
```

------------------------------------------------------------------------

## AC3 · Monthly Trend Chart {#sec-trend}

Dual-line chart showing total deliveries (KHIS) vs SHA claimed deliveries over time. The shaded area between the lines represents the **coverage gap**. Outliers in KHIS data are winsorised to the nearest IQR fence before plotting.

```{r}
#| label: chart-trend
#| fig-height: 5

# ── FRONTEND SPEC ─────────────────────────────────────────────
# Line 1 (blue,  solid):  Total deliveries — KHIS (D)
# Line 2 (green, solid):  SHA claimed deliveries (N)
# Shaded area between Line 1 and Line 2: coverage gap (neutral fill)
# Reference line at 80% coverage threshold (optional, secondary y-axis)
# Tooltip on hover: Month | Total (D) | SHA (N) | Proportion | Gap
# X-axis: Monthly (MMM YYYY)
# Y-axis left: count
# Mark winsorised months with a distinct point shape (triangle)
# ─────────────────────────────────────────────────────────────

p_trend <- ggplot(monthly, aes(x = period_month)) +

  # Gap fill between the two series
  geom_ribbon(
    aes(ymin = sha_total, ymax = khis_total),
    fill  = pal$gap,
    alpha = 0.6
  ) +

  # KHIS total deliveries
  geom_line(
    aes(y = khis_total, colour = "Total Deliveries (KHIS)"),
    linewidth = 1.2
  ) +
  geom_point(
    aes(y = khis_total, colour = "Total Deliveries (KHIS)",
        shape = as.factor(outliers > 0),
        text  = glue(
          "Month: {month_lbl}\n",
          "Total (KHIS): {format(khis_total, big.mark=',')}\n",
          "SHA Claimed:  {format(sha_total,  big.mark=',')}\n",
          "Proportion:   {round(proportion*100,1)}%\n",
          "Gap:          {format(gap, big.mark=',')}\n",
          "{if_else(outliers>0, '⚠ KHIS outlier(s) winsorised', '')}"
        )
    ),
    size = 3, stroke = 1.2, fill = "white"
  ) +

  # SHA claimed deliveries
  geom_line(
    aes(y = sha_total, colour = "SHA Claimed (N)"),
    linewidth = 1.2, linetype = "solid"
  ) +
  geom_point(
    aes(y = sha_total, colour = "SHA Claimed (N)"),
    size = 2.5
  ) +

  scale_colour_manual(
    values = c("Total Deliveries (KHIS)" = pal$primary,
               "SHA Claimed (N)"         = pal$accent),
    name = NULL
  ) +
  scale_shape_manual(
    values = c("FALSE" = 21, "TRUE" = 24),   # 24 = triangle → outlier month
    labels = c("Normal", "Outlier winsorised"),
    name   = "KHIS data"
  ) +
  scale_x_date(date_labels = "%b '%y", date_breaks = "1 month") +
  scale_y_continuous(
    labels = label_comma(),
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title    = "Monthly Deliveries: KHIS Total vs SHA Claimed",
    subtitle = "Shaded area = coverage gap · ▲ = KHIS outlier winsorised to IQR fence",
    x = NULL, y = "Delivery Count"
  )

ggplotly(p_trend, tooltip = "text") |>
  layout(legend = list(orientation = "h", x = 0, y = -0.18))
```

------------------------------------------------------------------------

## AC2 · County Comparison Table {#sec-table}

```{r}
#| label: county-table

# ── FRONTEND SPEC ─────────────────────────────────────────────
# Columns: County | KHIS Deliveries | SHA Claimed | Proportion
# Default sort: Proportion ASCENDING (lowest coverage first)
# Sortable by all columns (click header)
# Proportion column: colour-coded badge
#   🔴 Red   < 50 %
#   🟡 Amber 50 – 79 %
#   🟢 Green ≥ 80 %
# Counties with zero SHA claims MUST appear (left-join preserved)
# Table reacts to Date and County global filters
#
# ── BACKEND SPEC ──────────────────────────────────────────────
# GROUP BY county
# KHIS count : SUM of winsorised khis_count in filtered window
# SHA count  : SUM of sha_count in filtered window (0 if no claims)
# Proportion : sha_count / khis_count  (NULL-safe: 0 / 0 → 0)
# ─────────────────────────────────────────────────────────────

prop_badge <- function(val_pct, colour) {
  cls <- glue("badge-{colour}")
  glue('<span class="{cls}">{val_pct}%</span>')
}

county_tbl %>%
  mutate(prop_cell = map2_chr(prop_pct, prop_colour, prop_badge)) %>%
  select(
    County         = county,
    `KHIS Deliveries` = khis_count,
    `SHA Claimed`  = sha_count,
    `Proportion`   = prop_cell
  ) %>%
  gt() %>%
  tab_header(
    title    = md("**County-Level SHA Delivery Coverage**"),
    subtitle = glue("Period: {format(filter_date_min,'%b %Y')} – {format(filter_date_max,'%b %Y')} · Sorted by Proportion ↑")
  ) %>%
  fmt_number(columns = c(`KHIS Deliveries`, `SHA Claimed`),
             decimals = 0, use_seps = TRUE) %>%
  fmt_markdown(columns = `Proportion`) %>%
  cols_align(align = "right",  columns = c(`KHIS Deliveries`, `SHA Claimed`)) %>%
  cols_align(align = "center", columns = `Proportion`) %>%
  cols_align(align = "left",   columns = County) %>%
  tab_style(
    style     = list(cell_fill(color = "#1A56DB"),
                     cell_text(color = "white", weight = "bold")),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size                 = 13,
    heading.title.font.size         = 15,
    table.width                     = pct(100),
    table.border.top.color          = "transparent",
    table.border.bottom.color       = "transparent",
    row.striping.background_color   = "#F8F9FC",
    row.striping.include_table_body = TRUE
  ) %>%
  opt_row_striping() %>%
  tab_footnote(
    footnote  = md("🔴 <50% · 🟡 50–79% · 🟢 ≥80% · Counties with zero SHA claims are included."),
    locations = cells_title(groups = "subtitle")
  )
```

------------------------------------------------------------------------

## Appendix: Data Dictionary {#sec-appendix}

| Field | Description |
|---|---|
| `khis_count` | Monthly total deliveries from DHIS2 element `BDsWWkJHuce`, winsorised for outliers |
| `sha_count` | Monthly deliveries claimed under SHA-08-005, SHA-08-006, SHA-08-007 |
| `proportion` | `sha_count / khis_count` — SHA coverage rate (capped at 1.0) |
| `is_outlier` | `TRUE` if raw `khis_count` fell outside `[Q1 − 1.5·IQR, Q3 + 1.5·IQR]` |
| `period_month` | First day of the reporting month (join key between KHIS and SHA) |
| `county` | County name — must be harmonised between SHA and KHIS before joining |

**Intervention codes included**

| Code | Intervention |
|---|---|
| SHA-08-005 | Vaginal delivery |
| SHA-08-006 | Caesarean section |
| SHA-08-007 | Multiple births delivery |

**KHIS data element**

| Indicator | DHIS2 ID |
|---|---|
| Total deliveries | `BDsWWkJHuce` |

> **Note:** All figures are synthetic (generated with `set.seed(42)`).
> Replace the `df_sha` and `df_khis` data-generation blocks with real
> database / DHIS2 API queries. Apply the IQR winsorisation step
> server-side before joining.
