---
title: "Claims Turnaround Time (TAT) Analytics"
subtitle: "Submission → Payment Duration Analysis · 2024"
date: today
date-format: "D MMMM YYYY"
author: "Claims Analytics Team"
format:
  html:
    theme: [flatly, custom.scss]
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    toc-location: left
    number-sections: false
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    fig-width: 10
    fig-height: 5
    fig-dpi: 150
    embed-resources: true          # single self-contained HTML file
    page-layout: full
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false

# ── Install / load packages ──────────────────────────────────
pkgs <- c("tidyverse", "lubridate", "scales", "glue",
          "plotly", "gt", "gtExtras", "htmltools")
installed <- rownames(installed.packages())
to_install <- pkgs[!pkgs %in% installed]
if (length(to_install)) install.packages(to_install, quiet = TRUE)
invisible(lapply(pkgs, library, character.only = TRUE))

# ── Colour palette ────────────────────────────────────────────
pal <- list(
  bg      = "#F4F6FB",
  card    = "#FFFFFF",
  primary = "#1A56DB",
  accent  = "#7E3AF2",
  green   = "#0E9F6E",
  red     = "#E02424",
  amber   = "#D97706",
  text    = "#111928",
  muted   = "#6B7280",
  grid    = "#E5E7EB",
  border  = "#D1D5DB"
)

# ── ggplot2 base theme ────────────────────────────────────────
theme_tat <- function() {
  theme_minimal(base_family = "sans") +
    theme(
      plot.background  = element_rect(fill = pal$bg,   colour = NA),
      panel.background = element_rect(fill = pal$bg,   colour = NA),
      panel.grid.major = element_line(colour = pal$grid, linewidth = 0.4),
      panel.grid.minor = element_blank(),
      axis.text        = element_text(colour = pal$muted, size = 9),
      axis.title       = element_text(colour = pal$text,  size = 10),
      plot.title       = element_text(colour = pal$text,  size = 13,
                                      face = "bold", margin = margin(b = 4)),
      plot.subtitle    = element_text(colour = pal$muted, size = 9,
                                      margin = margin(b = 12)),
      legend.position  = "bottom",
      legend.text      = element_text(colour = pal$muted, size = 8),
      plot.margin      = margin(12, 16, 12, 16),
      strip.text       = element_text(colour = pal$text, face = "bold", size = 10)
    )
}
theme_set(theme_tat())
```

```{r}
#| label: data-generation
#| include: false

# ── Reproducible sample data ──────────────────────────────────
set.seed(42)
n_claims <- 1200

claims_raw <- tibble(
  claim_id     = paste0("CLM-", str_pad(seq_len(n_claims), 5, pad = "0")),
  created_date = sample(
    seq(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day"),
    n_claims, replace = TRUE
  )
) |>
  mutate(
    is_paid    = runif(n()) < 0.85,
    raw_tat    = pmax(1L, pmin(round(rlnorm(n(), meanlog = 3.0, sdlog = 0.55)), 120L)),
    paid_date  = if_else(is_paid, created_date + raw_tat, as.Date(NA)),
    tat_days   = as.numeric(paid_date - created_date),
    created_month = floor_date(created_date, "month"),
    paid_month    = floor_date(paid_date,    "month")
  )

paid_claims <- claims_raw |> filter(is_paid)

# ── Reference months ─────────────────────────────────────────
cur_month  <- as.Date("2024-11-01")
prev_month <- cur_month %m-% months(1)

cur_paid  <- paid_claims |> filter(paid_month == cur_month)
prev_paid <- paid_claims |> filter(paid_month == prev_month)
cur_all   <- claims_raw  |> filter(created_month == cur_month)

# ── Metric helpers ────────────────────────────────────────────
metrics <- function(df) {
  list(
    n_paid     = nrow(df),
    min_tat    = min(df$tat_days),
    median_tat = median(df$tat_days),
    mean_tat   = mean(df$tat_days),
    max_tat    = max(df$tat_days),
    pct_30     = mean(df$tat_days <= 30) * 100
  )
}

m_cur  <- metrics(cur_paid)
m_prev <- metrics(prev_paid)

delta <- list(
  min_tat    = m_cur$min_tat    - m_prev$min_tat,
  median_tat = m_cur$median_tat - m_prev$median_tat,
  max_tat    = m_cur$max_tat    - m_prev$max_tat,
  pct_30     = m_cur$pct_30     - m_prev$pct_30
)

fmt_delta <- function(x, unit = "days", good_dir = -1) {
  # good_dir: -1 = lower is better, +1 = higher is better
  colour <- if (x * good_dir < 0) pal$green else if (x * good_dir > 0) pal$red else pal$muted
  sign   <- if (x >= 0) "▲ +" else "▼ "
  label  <- glue("{sign}{round(abs(x), 1)} {unit} vs prev month")
  list(label = label, colour = colour)
}

d_min    <- fmt_delta(delta$min_tat)
d_median <- fmt_delta(delta$median_tat)
d_max    <- fmt_delta(delta$max_tat)
d_pct    <- fmt_delta(delta$pct_30, unit = "pp", good_dir = 1)
```

```{r}
#| label: card-css
#| echo: false
#| results: asis

# Inline CSS for metric cards (avoids needing custom.scss for the cards)
cat('
<style>
  .card-row   { display: flex; gap: 14px; flex-wrap: wrap; margin-bottom: 28px; }
  .metric-card {
    flex: 1 1 140px;
    background: #fff;
    border: 1px solid #E5E7EB;
    border-radius: 10px;
    padding: 18px 20px 14px;
    box-shadow: 0 1px 4px rgba(0,0,0,.06);
  }
  .metric-card .card-label  { font-size: 11px; color: #6B7280; text-transform: uppercase;
                               letter-spacing: .06em; margin-bottom: 6px; }
  .metric-card .card-value  { font-size: 28px; font-weight: 700; color: #111928;
                               line-height: 1.1; }
  .metric-card .card-delta  { font-size: 11px; margin-top: 8px; font-weight: 500; }
  .metric-card .card-sub    { font-size: 10px; color: #9CA3AF; margin-top: 2px; }
  .section-label {
    font-size: 11px; text-transform: uppercase; letter-spacing: .08em;
    color: #6B7280; margin-bottom: 10px; font-weight: 600;
  }
</style>
')
```

## AC1 · Summary Metrics {#sec-metrics}

::: section-label
November 2024 — Paid Date − Created Date
:::

```{r}
#| label: metric-cards
#| echo: false
#| results: asis

card <- function(label, value, delta_label = NULL, delta_colour = NULL, sub = NULL) {
  delta_html <- if (!is.null(delta_label))
    glue('<div class="card-delta" style="color:{delta_colour}">{delta_label}</div>')
  else ""
  sub_html <- if (!is.null(sub))
    glue('<div class="card-sub">{sub}</div>')
  else ""
  glue('
  <div class="metric-card">
    <div class="card-label">{label}</div>
    <div class="card-value">{value}</div>
    {delta_html}
    {sub_html}
  </div>')
}

cards_html <- paste0(
  '<div class="card-row">',
  card("Total Claims",       format(nrow(cur_all),    big.mark = ","),
       sub = "Created in Nov 2024"),
  card("Paid Claims",        format(m_cur$n_paid,     big.mark = ","),
       sub = glue("{round(m_cur$n_paid/nrow(cur_all)*100,1)}% of total")),
  card("Min TAT",            glue("{m_cur$min_tat} days"),
       d_min$label,    d_min$colour),
  card("Median TAT",         glue("{round(m_cur$median_tat,1)} days"),
       d_median$label, d_median$colour),
  card("Max TAT",            glue("{m_cur$max_tat} days"),
       d_max$label,    d_max$colour),
  card("% Paid ≤ 30 days",  glue("{round(m_cur$pct_30,1)}%"),
       d_pct$label,    d_pct$colour),
  '</div>'
)
cat(cards_html)
```

------------------------------------------------------------------------

## AC2 · TAT Trend Over Time {#sec-trend}

The charts below show average TAT for claims that **completed** (received a paid date) within each period. Use the toggle to switch between weekly and monthly granularity.

```{r}
#| label: trend-data

# Weekly
weekly_tat <- paid_claims |>
  mutate(period = floor_date(paid_date, "week", week_start = 1)) |>
  group_by(period) |>
  summarise(
    avg_tat = mean(tat_days),
    med_tat = median(tat_days),
    n       = n(),
    .groups = "drop"
  ) |>
  filter(n >= 5) |>
  mutate(granularity = "Weekly")

# Monthly
monthly_tat <- paid_claims |>
  mutate(period = floor_date(paid_date, "month")) |>
  group_by(period) |>
  summarise(
    avg_tat = mean(tat_days),
    med_tat = median(tat_days),
    n       = n(),
    .groups = "drop"
  ) |>
  mutate(granularity = "Monthly")
```

::: panel-tabset
### 📅 Weekly

```{r}
#| label: chart-weekly
#| fig-height: 4.5

p_week <- ggplot(weekly_tat, aes(x = period)) +
  geom_ribbon(aes(ymin = avg_tat - 4, ymax = avg_tat + 4),
              fill = pal$primary, alpha = 0.10) +
  geom_line(aes(y = avg_tat, colour = "Mean TAT"),  linewidth = 1.1) +
  geom_line(aes(y = med_tat, colour = "Median TAT"), linewidth = 0.8, linetype = "dashed") +
  geom_point(aes(y = avg_tat, colour = "Mean TAT",
                 text = glue("Week: {format(period,'%d %b %Y')}\nMean TAT: {round(avg_tat,1)} days\nClaims: {n}")),
             size = 2, shape = 21, fill = "white", stroke = 1.5) +
  geom_hline(yintercept = 30, linetype = "dotted", colour = pal$red,
             linewidth = 0.7) +
  annotate("text", x = min(weekly_tat$period) + 3, y = 31.8,
           label = "30-day target", hjust = 0, size = 2.8, colour = pal$red) +
  scale_colour_manual(values = c("Mean TAT" = pal$primary, "Median TAT" = pal$accent),
                      name = NULL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "4 weeks") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .12)),
                     labels = \(x) paste0(x, "d")) +
  labs(title    = "Average TAT by Week",
       subtitle = "Jan – Dec 2024 · Plotted on paid date · Shaded band = ±4 days",
       x = NULL, y = "Avg TAT (days)")

ggplotly(p_week, tooltip = "text") |>
  layout(legend = list(orientation = "h", x = 0, y = -0.15))
```

### 📆 Monthly

```{r}
#| label: chart-monthly
#| fig-height: 4.5

p_month <- ggplot(monthly_tat, aes(x = period)) +
  geom_col(aes(y = avg_tat,
               text = glue("{format(period,'%b %Y')}\nMean: {round(avg_tat,1)}d\nMedian: {round(med_tat,1)}d\nClaims: {n}")),
           fill = pal$primary, alpha = 0.75, width = 22) +
  geom_line(aes(y = avg_tat),  colour = pal$primary, linewidth = 1.1) +
  geom_line(aes(y = med_tat),  colour = pal$accent,  linewidth = 0.9, linetype = "dashed") +
  geom_point(aes(y = avg_tat), colour = pal$primary, size = 3) +
  geom_point(aes(y = med_tat), colour = pal$accent,  size = 2.5, shape = 21,
             fill = "white", stroke = 1.2) +
  geom_text(aes(y = avg_tat, label = round(avg_tat, 1)),
            vjust = -0.9, size = 2.7, colour = pal$text) +
  geom_hline(yintercept = 30, linetype = "dotted", colour = pal$red, linewidth = 0.7) +
  annotate("text", x = min(monthly_tat$period) + 5, y = 31.8,
           label = "30-day target", hjust = 0, size = 2.8, colour = pal$red) +
  scale_x_date(date_labels = "%b '%y", date_breaks = "1 month") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .15)),
                     labels = \(x) paste0(x, "d")) +
  labs(title    = "Average TAT by Month",
       subtitle = "Jan – Dec 2024 · Bars = mean · Dashed line = median · Label = mean days",
       x = NULL, y = "Avg TAT (days)")

ggplotly(p_month, tooltip = "text") |>
  layout(showlegend = FALSE)
```
:::

------------------------------------------------------------------------

## TAT Distribution {#sec-dist}

```{r}
#| label: chart-distribution
#| fig-height: 4

p_dist <- paid_claims |>
  mutate(month_lbl = format(paid_month, "%b")) |>
  filter(paid_month >= as.Date("2024-07-01")) |>   # last 6 months for readability
  ggplot(aes(x = tat_days, y = after_stat(density))) +
  geom_histogram(fill = pal$primary, alpha = 0.7, binwidth = 5, boundary = 0) +
  geom_density(colour = pal$accent, linewidth = 0.9) +
  geom_vline(xintercept = 30, colour = pal$red, linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = 31, y = Inf, label = "30d", hjust = 0, vjust = 1.5,
           size = 3, colour = pal$red) +
  facet_wrap(~month_lbl, nrow = 2) +
  scale_x_continuous(limits = c(0, 100), breaks = c(0, 30, 60, 90)) +
  labs(title    = "TAT Distribution by Month (Jul – Dec 2024)",
       subtitle = "Histogram + density curve · Dashed = 30-day target",
       x = "TAT (days)", y = "Density")

p_dist
```

------------------------------------------------------------------------

## Monthly Summary Table {#sec-table}

```{r}
#| label: summary-table

table_data <- paid_claims |>
  mutate(month = floor_date(paid_date, "month")) |>
  group_by(Month = month) |>
  summarise(
    `Claims Paid`  = n(),
    `Min TAT`      = min(tat_days),
    `P25 TAT`      = quantile(tat_days, 0.25),
    `Median TAT`   = round(median(tat_days), 1),
    `Mean TAT`     = round(mean(tat_days), 1),
    `P75 TAT`      = quantile(tat_days, 0.75),
    `Max TAT`      = max(tat_days),
    `% ≤ 30 days`  = round(mean(tat_days <= 30) * 100, 1),
    .groups = "drop"
  ) |>
  mutate(Month = format(Month, "%b %Y"))

table_data |>
  gt() |>
  tab_header(
    title    = md("**Monthly TAT Summary — 2024**"),
    subtitle = "TAT = Paid Date − Created Date"
  ) |>
  tab_spanner(
    label   = "Turnaround Time (days)",
    columns = c(`Min TAT`, `P25 TAT`, `Median TAT`, `Mean TAT`, `P75 TAT`, `Max TAT`)
  ) |>
  fmt_number(columns = `Claims Paid`,  decimals = 0, use_seps = TRUE) |>
  fmt_number(columns = c(`Min TAT`, `P25 TAT`, `Median TAT`,
                         `Mean TAT`, `P75 TAT`, `Max TAT`),
             decimals = 1) |>
  fmt_number(columns = `% ≤ 30 days`, decimals = 1, pattern = "{x}%") |>
  cols_align(align = "center", columns = -Month) |>
  cols_align(align = "left",   columns = Month) |>
  data_color(
    columns = `Median TAT`,
    method  = "numeric",
    palette = c("#0E9F6E", "#FFFBEB", "#E02424"),
    domain  = c(15, 40)
  ) |>
  data_color(
    columns = `% ≤ 30 days`,
    method  = "numeric",
    palette = c("#E02424", "#FFFBEB", "#0E9F6E"),
    domain  = c(50, 90)
  ) |>
  gtExtras::gt_highlight_rows(
    rows      = Month == "Nov 2024",
    fill      = "#EFF6FF",
    font_weight = "bold"
  ) |>
  tab_style(
    style     = list(cell_fill(color = "#1A56DB"),
                     cell_text(color = "white", weight = "bold")),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = list(cell_fill(color = "#1A56DB"),
                     cell_text(color = "white", weight = "bold")),
    locations = cells_column_spanners()
  ) |>
  tab_options(
    table.font.size               = 13,
    heading.title.font.size       = 15,
    table.width                   = pct(100),
    table.border.top.color        = "transparent",
    table.border.bottom.color     = "transparent",
    row.striping.background_color = "#F8F9FC",
    row.striping.include_table_body = TRUE
  ) |>
  opt_row_striping() |>
  tab_footnote(
    footnote  = "Highlighted row = current reporting month (Nov 2024)",
    locations = cells_title(groups = "subtitle")
  )
```

------------------------------------------------------------------------

## Appendix: Data Dictionary {#sec-appendix}

| Field          | Description                                 |
|----------------|---------------------------------------------|
| `claim_id`     | Unique claim identifier                     |
| `created_date` | Date the claim was submitted (TAT start)    |
| `paid_date`    | Date the claim was paid (TAT end)           |
| `tat_days`     | `paid_date − created_date` in calendar days |
| `is_paid`      | Boolean — whether a paid date exists        |

> **Note:** Sample data is randomly generated with `set.seed(42)`. Replace the data generation block with your actual claims extract.
